---
title: "Mahalanobis Distance"
output: html_notebook
---

# Introduction

This notebook is a snippet about Mahalanobis Distance to be used to supplement the material in other notebooks in this directory.

I populated this notebook by following the material [here](https://www.analyticsindiamag.com/understanding-mahalanobis-distance-and-its-use-cases/). In an effort to better understand the material, I will be implementing a formula which can make this calculation from scratch. In reality, there are built-in functions which achieves this effect, which would be used in production code.

# What is it?

Mahalanobis distance (MD) measures the relative distance between two variables with respect to the centroid.

# Equation

Observataions:
$$x = (x_1,x_2,x_3,...x_N)^T$$

Means:
$$\mu=(\mu_1,\mu_2,\mu_3,...\mu_N)^T$$

Covariance Matrix $$S$$

Mahalanobis Distance Formula
$$MD(x)=\sqrt{(x-\mu)^TS^{-1}(x-\mu)}$$

# As an R Function

```{r}
m1 <- matrix(sample(0:9, 12, replace = T), ncol=4)
m2 <- matrix(sample(0:9, 12, replace = T), ncol=3)

cov(m1)
cor(m1)

mahalanobis_distance <- function(x, means, covariance){
  
}

det(m1)

library(matlib)

mahalanobis_distance(c(2, 4, 3), c())

```

# Testing MD

```{r}

q1 <- c(0, 1, 0, 0, 0, 1, 0.2)
q2 <- c(1, 1, 1, 1, 1, 1, 0)
q3 <- c(1, 0, 1, 0, 1, 1, 0.5)

mat <- cbind(q1, q2, q3)

md <- function(x){
  if(is.data.frame(x)){
    x <- as.matrix(x)
  }
  if(is.matrix(x)){
    mahalanobis(x, colMeans(x), cov(x))
  } else {
    warning('x must be a matrix or a data.frame')
  }
}


md(mat)
mahalanobis(mat, colMeans(mat), cov(mat))

qchisq(0.99, 50)


```




















